\documentclass[acmsmall,nonacm]{acmart}

% acmart already provides amsmath/amsthm/amssymb, booktabs, graphicx,
% xcolor, array, and hyperref; add only what it does not.
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}
\usepackage{makecell}
\usepackage{float}
\setlength{\emergencystretch}{3em} % absorb overfull lines from unbreakable inline code

\settopmatter{printacmref=false}   % preprint: suppress the ACM reference block
\raggedbottom                      % figure-heavy: let slack fall to page bottom,
                                   % not stretch above pinned [H] figures

%% Theorem environments (acmart styles: `invariant` italic, `definition` upright)
\theoremstyle{acmplain}
\newtheorem{invariant}{Invariant}
\theoremstyle{acmdefinition}
\newtheorem{definition}{Definition}
%% Inline code
\newcommand{\code}[1]{\texttt{#1}}

%% Java listings
\definecolor{codegray}{gray}{0.45}
\definecolor{codeblue}{rgb}{0.13,0.29,0.53}
\definecolor{codebg}{rgb}{0.97,0.97,0.97}
\lstdefinestyle{java}{
  language=Java,
  basicstyle=\small\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  keywordstyle=\color{codeblue}\bfseries,
  commentstyle=\color{codegray}\itshape,
  stringstyle=\color{codegray},
  backgroundcolor=\color{codebg},
  frame=single,
  framesep=4pt,
  rulecolor=\color{black!15},
  xleftmargin=6pt,
  xrightmargin=6pt,
  breaklines=true,
  breakatwhitespace=true,
  showstringspaces=false,
  tabsize=2,
  numbers=none,
  morekeywords={abstract,implements,extends,interface,class,new,return,
                public,private,protected,static,final,void,String,boolean,
                int,Object,Override,System,out,println},
}
\lstset{style=java}

%% Kotlin listings
\lstdefinestyle{kotlin}{
  language=,
  basicstyle=\small\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  keywordstyle=\color{codeblue}\bfseries,
  commentstyle=\color{codegray}\itshape,
  stringstyle=\color{codegray},
  backgroundcolor=\color{codebg},
  frame=single,
  framesep=4pt,
  rulecolor=\color{black!15},
  xleftmargin=6pt,
  xrightmargin=6pt,
  breaklines=true,
  breakatwhitespace=true,
  showstringspaces=false,
  tabsize=2,
  numbers=none,
  morekeywords={fun,val,var,class,interface,override,by,abstract,open,
                private,public,protected,String,Unit,Any,return,if,else},
}


\begin{document}

\title{Interface-Scoped Dispatch:\texorpdfstring{\\}{ }Independent Components with O(1) Self-Calls}

\author{R. Scott McKinney}
\affiliation{\institution{Manifold Systems}\country{USA}}
\email{scott@manifold.systems}
\renewcommand{\shortauthors}{McKinney}

\begin{abstract}
Object composition under static typing forces a trade: independent
components, or self-calls through the composite. Forwarding-based constructs compose independent components, but a component's self-calls do not dispatch through the composite. Trait/mixin composition preserves self-call dispatch at the expense of independent
components. Neither delivers both. Achieving both at the cost of ordinary
dispatch is the open problem.

We present \emph{Parts}, realized as a compiler plugin for Java~8--26.
A \emph{part} is an independent component, supplied as a runtime object at
construction, whose self-calls route through the composite; the enabling
mechanism, \emph{interface-scoped dispatch} (ISD), assigns composite identity
per interface rather than per object. Self-calls route
through the correct composite in arbitrary DAG compositions. The dispatch cost is O(1), empirically equivalent to a vtable call.
\end{abstract}

\maketitle

%% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
%% ─────────────────────────────────────────────────────────────────────────────

Object composition is the standard prescription for avoiding implementation
inheritance~\cite{gamma1994}. Two properties make it a genuine replacement:
\emph{independent components} (unrestricted late-bound objects fixed at
construction) and \emph{open recursion} (self-calls through the composite).
Static typing is what makes the two hard to combine: dynamically-typed object
models have had both for decades, yet to our knowledge no available statically-typed
language provides them together---and none at the cost of ordinary dispatch.

Forwarding-based composition (Kotlin \code{by}~\cite{kotlin}, Lombok
\code{@Delegate}~\cite{lombok}) supplies independent components but does not
support open recursion: a component's self-calls resolve within the component
because the composite is never on the dispatch path. This is
the \emph{self} problem~\cite{lieberman1986,self1987}.

Trait composition (Scala \code{with}~\cite{odersky2005},
mixins~\cite{bracha1990}) inverts the trade. Self-calls dispatch through
\code{this} in the linearized class, so composite overrides are visible. But
traits are fused into the class hierarchy at compile time, not supplied as
runtime objects. Components cannot be constructed independently or supplied externally;
composition is static.

\emph{Parts} provides both. \code{@part} marks a class as a \emph{part}, an independent object supplied at construction. \code{@link} marks a field through which a composite
delegates one or more interfaces to a part. When a composite is constructed, a
compiler plugin wires composite identity into each part, so self-calls in
the part dispatch through the composite. A link field may be assigned any
implementation of its interface: a non-part object is driven exclusively by forwarding; a part does what forwarding structurally cannot---route
its self-calls through the composite. Parts thus subsumes forwarding rather than replacing
it. The enabling mechanism is
\emph{interface-scoped dispatch} (ISD): assigning composite identity per
interface rather than per object enables open recursion in all Parts compositions.

The core contributions are:

\begin{itemize}
  \item \textbf{Parts}: a model for object composition in which
        components are independent runtime objects, self-calls dispatch through
        the composite, and interface contracts are statically enforced.
        The model is defined over standard interface hierarchies with a central
        dispatch invariant (Invariant~\ref{prop:dispatch}) and
        composite-scoped encapsulation (\code{@internal}).
  \item \textbf{Interface-scoped dispatch (ISD)}: the mechanism that assigns
        composite identity per interface rather than per object, enabling
        open recursion in Parts compositions, including partial
        delegation (where a composite claims a proper subset of a part's
        interfaces) and arbitrary DAG compositions.
  \item \textbf{An implementation}: a compiler plugin for
        Java 8--26 realizing the model at O(1) dispatch cost, with open-source
        release~\cite{mckinney2026} and IDE integration.
\end{itemize}

%% ─────────────────────────────────────────────────────────────────────────────
\section{Overview}
%% ─────────────────────────────────────────────────────────────────────────────

Consider a minimal example that shows \code{@part} and \code{@link} working in a concrete single-rooted composite.

\begin{lstlisting}
interface Formatter {
    String prefix();
    String format(String msg);
}

@part class LogFormatter implements Formatter {
    public String prefix() { return "INFO"; }
    public String format(String msg) { return "[" + prefix() + "] " + msg; }
}

class AppFormatter implements Formatter {
    @link Formatter fmt;
    AppFormatter(Formatter fmt) { this.fmt = fmt; }
    @Override public String prefix() { return "APP"; }
}

new AppFormatter(new LogFormatter()).format("started"); // "[APP] started"
\end{lstlisting}

When \code{AppFormatter} is constructed, the compiler wires
\code{LogFormatter}'s composite identity for \code{Formatter} to
\code{AppFormatter}. A call to \code{format("started")} dispatches to
\code{LogFormatter.format()}, and the self-call \code{prefix()} within
\code{format()} dispatches back through \code{AppFormatter}, invoking
\code{AppFormatter.prefix()}, not \code{LogFormatter.prefix()}.

Because the link field accepts any \code{Formatter} at construction
time, the part is a real late-bound object, not fused into the class
hierarchy. A different \code{Formatter} implementation can be substituted
without changing \code{AppFormatter}.

Interface contracts are enforced at compile time: the link field has
type \code{Formatter}, and the compiler verifies that \code{AppFormatter}
implements the delegated interface. Part identity may not escape the
composite dispatch path: passing or assigning \code{this} as the concrete
part type is a compile error; interface types are required.

This example is single-rooted: \code{AppFormatter} is the sole composite
for \code{LogFormatter}, and one composite identity suffices. The following
section exposes the limit of the single-identity model and the per-interface
generalization that resolves it.

%% ─────────────────────────────────────────────────────────────────────────────
\section{The Model}
\label{sec:model}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection*{Motivating example}

The following example shows a part whose interfaces are rooted at different
composites. \code{Track} implements both \code{Streamable} and
\code{Classifiable}; in the composed graph, \code{Streamable} is rooted at
\code{Playlist} and \code{Classifiable} at \code{Library}. From one
\code{Track} instance, \code{codec()} must route through \code{Playlist} and
\code{genre()} through \code{Library}; a single composite identity cannot serve
both. Assigning a separate composite identity per interface resolves this.
Figures~\ref{fig:dag-diagram} and~\ref{fig:dag-listing} show the structure.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  cls/.style={
    draw, rectangle, rounded corners=3pt,
    minimum width=5.5cm, inner sep=8pt,
    align=center, font=\small
  },
  lnk/.style={->, >=Stealth, semithick},
]
\node[cls] (wp)
  {\textbf{Track}\\[3pt]
   \textit{implements Streamable, Classifiable}};

\node[cls, below=2cm of wp] (lw)
  {\textbf{Playlist}\\[3pt]
   \textit{implements Streamable, Classifiable}};

\node[cls, below=2cm of lw] (app)
  {\textbf{Library}\\[3pt]
   \textit{implements Classifiable}};

%% Playlist @link -> Track
\draw[lnk] ([xshift=-12pt]lw.north)
  -- node[left,  font=\scriptsize\ttfamily] {@link Streamable}
  ([xshift=-12pt]wp.south);
\draw[lnk] ([xshift=12pt]lw.north)
  -- node[right, font=\scriptsize\ttfamily] {@link Classifiable}
  ([xshift=12pt]wp.south);

%% Library @link -> Playlist
\draw[lnk] (app.north)
  -- node[right, font=\scriptsize\ttfamily] {@link Classifiable}
  (lw.south);

%% Dispatch root annotations
\node[font=\scriptsize\itshape, text=gray, right=0.5cm of lw, anchor=west]
  {Streamable root};
\node[font=\scriptsize\itshape, text=gray, right=0.5cm of app, anchor=west]
  {Classifiable root};
\end{tikzpicture}
\caption{DAG composition structure. Arrows show link fields labeled
by the delegated interface. \code{Playlist} is the root for
\code{Streamable} dispatch; \code{Library} is the root for \code{Classifiable} dispatch.}
\label{fig:dag-diagram}
\Description{A three-class DAG composition. Track implements Streamable and
Classifiable. Playlist links both interfaces of Track and is the Streamable
root. Library links the Classifiable interface of Playlist and is the
Classifiable root.}
\end{figure}

\begin{figure}[H]
\begin{lstlisting}
interface Streamable {
    String codec();
    void stream(String audio);
}

interface Classifiable {
    String genre();
    String classify(String title);
}

@part class Track implements Streamable, Classifiable {
  public String codec()  { return "wav"; }
  public String genre()  { return "unknown"; }
  public void stream(String audio) {
    System.out.println("[" + codec() + "] " + audio);
  }
  public String classify(String title) {
    stream("indexing: " + title);   // Streamable dispatch
    return genre() + ":" + title;   // Classifiable dispatch
  }
}

@part class Playlist implements Streamable, Classifiable {
  @link Streamable   streamable;
  @link Classifiable classifiable;
  Playlist(Track t) { this.streamable = t; this.classifiable = t; }
  @Override public String codec() { return "mp3"; }
}

class Library implements Classifiable {
  @link Classifiable classifiable;
  Library(Classifiable c) { this.classifiable = c; }
  @Override public String genre() { return "jazz"; }
}

Playlist playlist = new Playlist(new Track());
Library library   = new Library(playlist);
String result = library.classify("Autumn Leaves");
out.println(result);
// Streamable dispatch rooted at Playlist  ->  codec()  -> "mp3"
// Classifiable dispatch rooted at Library ->  genre()  -> "jazz"
// prints:  "[mp3] indexing: Autumn Leaves"
// returns: "jazz:Autumn Leaves"
\end{lstlisting}
\caption{DAG composition with multiple composite roots per part.
\code{Track} carries distinct composite identities for \code{Streamable}
(rooted at \code{Playlist}) and \code{Classifiable} (rooted at \code{Library}).}
\label{fig:dag-listing}
\Description{Java source listing of the Track, Playlist, and Library
composition, showing distinct composite roots per interface.}
\end{figure}

\subsection*{Definitions}

We define Parts over a standard typed object model where
classes implement interfaces and method dispatch is defined by the receiver's
runtime type.

\begin{definition}[Part]
A \emph{part} is a class annotated with \code{@part}. Parts implement one
or more interfaces. A part may also be a composite.
\end{definition}

\begin{definition}[Link]
A \emph{link} is a field annotated with \code{@link}. Its declared type is an
interface or an abstract part class. A link claims the interfaces common
to its field type's interface hierarchy and the enclosing composite's. The
plugin enforces \code{private} and \code{final} on link fields; explicit
modifiers are redundant.
\end{definition}

\begin{definition}[Composite]
A \emph{composite} is a class containing one or more link fields. A
composite implements all interfaces claimed by its links.
\end{definition}

\begin{definition}[Delegation surface]
The \emph{delegation surface} of a part is its interface hierarchy,
excluding any interfaces marked \code{@internal} in its \code{implements} clause.
\end{definition}

\begin{definition}[Composite identity]
When a part is wired to a composite, it receives a \emph{composite
identity}: a reference to the composite object. Self-calls within the
part (calls to methods on the part's delegation surface) dispatch through
a composite identity rather than \code{this}.
\end{definition}

\begin{definition}[Interface-scoped dispatch]
\label{def:isd}
\emph{Interface-scoped dispatch} is the dispatch mechanism by which the
receiver for a self-call is determined by the interface of the
called method, not by \code{this}. Each
interface in a part's delegation surface is assigned an independent
composite identity; self-calls route through the composite for that
interface.
\end{definition}

\begin{definition}[Transitive propagation]
A part may itself contain link fields. When such a part is wired to
an outer composite, composite identity propagates transitively: the outer composite
becomes the dispatch root for each interface it claims throughout the delegation chain.
Interfaces not claimed by the outer composite retain the dispatch roots established
by inner composites. Propagation traverses part links only: it terminates at
any non-part link, which is a forwarding leaf whose self-calls remain
local to the delegate. A composition may freely mix parts and non-part delegates: each
part routes, each non-part forwards, with no interaction between them.
\end{definition}

\begin{definition}[Diamond resolution]
\label{def:diamond}
When two links in a composite share a common interface, exactly one must declare
\code{@link(share=T.class)} for the shared interface. That link is
authoritative; the other delegates only its non-shared interfaces, and the
shared interface is excluded from propagation when the non-authoritative link is
wired. Unresolved diamonds are compile errors.
\end{definition}

Two practical extensions, composite-scoped encapsulation (\code{@internal})
and abstract parts with deferred linkage (\code{asLink()}), are defined
and implemented (Section~\ref{sec:impl}).

\paragraph{Dispatch rule.}
Traditional object-oriented dispatch resolves a call through the receiver's
object identity:
\[ \text{call } m() \text{ on } o \;\longrightarrow\; \mathit{vtable}(o,\, m) \]
Interface-scoped dispatch replaces the single object identity with a
per-interface composite identity. A self-call $m()$ in part $P$ where
$m \in I$ resolves through the composite identity for $I$:
\[ \text{self-call } m() \in I \text{ in } P \;\longrightarrow\;
   \$\mathit{selves}_P[I].m() \]
For self-calls within a part's methods, the dispatch triple is
\emph{(part, interface, method)} rather than the conventional
\emph{(object, method)}.
Figure~\ref{fig:dispatch-comparison} contrasts this per-interface routing with a single shared composite identity.
This is \emph{open recursion}: a method's self-calls late-bind to the
most-derived receiver. Here the receiver resolves from the interface of the self-call. Forwarding is the closed case: a
component's self-calls stay in the component, and the composite never enters
the recursion.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  >=Stealth,
  cbox/.style={draw, rectangle, minimum width=2.6cm, minimum height=0.8cm,
               inner sep=4pt, align=center, font=\small},
  svbox/.style={draw=black!45, fill=gray!15, rectangle, minimum width=2.6cm,
                inner sep=2pt, align=center, font=\scriptsize\ttfamily},
  sbox/.style={draw=black!45, fill=gray!15, rectangle, minimum width=2.6cm,
               inner sep=2pt, align=center, font=\scriptsize\ttfamily},
  darr/.style={->, black!45, semithick},
  lbl/.style={font=\scriptsize\itshape, text=black!55},
]

%%====================================================================
%% LEFT PANEL -- Object Scoped
%%====================================================================
\node[font=\bfseries] at (-3.0, 2.0) {Object Scoped};
\node[lbl, text=black!75] at (-3.0, 1.5)
  {\textit{self-call} $m \in I_1$ inside $P$};

\node[cbox] (PL) at (-3.0, 0.8) {$P$ \ impl $I_1$,\,$I_2$};
\node[sbox, below=0pt of PL, minimum height=0.74cm] (BL) {\texttt{\$self}};
\node[cbox] (C1L) at (-3.0, -1.7) {$C_1$ \ impl $I_1$,\,$I_2$};
\node[cbox] (C2L) at (-3.0, -3.6) {$C_2$ \ impl $I_2$};

\draw[darr] (C1L.north) -- (BL.south)
  node[midway, right, xshift=3pt, lbl] {delegate $I_1$,\,$I_2$};
\draw[darr] (C2L.north) -- (C1L.south)
  node[midway, right, xshift=3pt, lbl] {delegate $I_2$};

\draw[{Circle[fill=red!80!black, scale=0.7]}->, red!80!black, thick, dashed]
  (BL.west)
  .. controls ($(BL.west)+(-1.1, 0)$) and ($(C2L.west)+(-1.1, 0)$) ..
  (C2L.west)
  node[pos=0.5, anchor=east, xshift=-4pt, lbl, align=right,
       text=red!70!black]
    {dispatch $I_1$ \ $\times$};

\node[font=\small, text=red!80!black, below=0.2cm of C2L, align=center]
  {wrong dispatch target for $I_1$};

%%====================================================================
%% DIVIDER
%%====================================================================
\draw[black!20, dashed] (0, 2.4) -- (0, -4.8);

%%====================================================================
%% RIGHT PANEL -- Interface Scoped
%%====================================================================
\node[font=\bfseries] at (3.0, 2.0) {Interface Scoped};
\node[lbl, text=black!75] at (3.0, 1.5)
  {\textit{self-call} $m \in I_1$ inside $P$};

\node[cbox] (PR) at (3.0, 0.8) {$P$ \ impl $I_1$,\,$I_2$};
\node[svbox, below=0pt of PR, minimum height=0.74cm] (BR)
  {\texttt{\$selves[$I_1$]}\\[1pt]\texttt{\$selves[$I_2$]}};
\node[cbox] (C1R) at (3.0, -1.7) {$C_1$ \ impl $I_1$,\,$I_2$};
\node[cbox] (C2R) at (3.0, -3.6) {$C_2$ \ impl $I_2$};

\draw[darr] (C1R.north) -- (BR.south)
  node[midway, left, xshift=-3pt, lbl] {delegate $I_1$,\,$I_2$};
\draw[darr] (C2R.north) -- (C1R.south)
  node[midway, left, xshift=-3pt, lbl] {delegate $I_2$};

\coordinate (slotI1) at ($(BR.north east)!0.27!(BR.south east)$);
\coordinate (slotI2) at ($(BR.north east)!0.73!(BR.south east)$);

\draw[{Circle[fill=black, scale=0.7]}->, black, thick]
  (slotI1)
  .. controls ($(slotI1)+(0.85, 0)$) and ($(C1R.east)+(0.85, 0)$) ..
  (C1R.east)
  node[pos=0.5, anchor=west, xshift=4pt, lbl, text=black!80]
    {dispatch $I_1$ \ $\checkmark$};

\draw[{Circle[fill=black!30, scale=0.7]}->, black!30, semithick]
  (slotI2)
  .. controls ($(slotI2)+(1.3, 0)$) and ($(C2R.east)+(1.3, 0)$) ..
  (C2R.east)
  node[pos=0.5, anchor=west, xshift=4pt, lbl, text=black!35]
    {dispatch $I_2$};

\end{tikzpicture}
\caption{Object-scoped vs.\ interface-scoped dispatch. In both panels $C_2$
(impl~$I_2$) delegates $I_2$ to $C_1$ (impl~$I_1$,\,$I_2$), which delegates
both interfaces to part $P$.
\emph{Left:} a self-call $m() \in I_1$ inside $P$ routes through
\texttt{\$self\,=\,$C_2$}, the wrong composite for $I_1$.
\emph{Right:} \texttt{\$selves[$I_1$]} routes to $C_1$;
\texttt{\$selves[$I_2$]} routes to $C_2$; each self-call reaches its
correct composite root.}
\label{fig:dispatch-comparison}
\Description{Side-by-side comparison of object-scoped versus interface-scoped
dispatch. Left panel shows a single identity slot routing an $I_1$ self-call
to the wrong composite. Right panel shows per-interface slots routing each
self-call to its correct composite root.}
\end{figure}

\begin{invariant}
\label{prop:dispatch}
For any part $P$ in a delegation chain for interface $I$ rooted at
composite $C$: a self-call \code{this.m()} within $P$ where $m \in I$
dispatches to $C$.
\end{invariant}

This invariant presumes each interface has a single composite root. That
precondition holds for every composition the model admits except the multi-root
contention case of Section~\ref{sec:limitations}, which is not statically
enforced; where it is violated, $C$ is not uniquely defined and the guarantee
does not hold.

The invariant is established by construction. When composite $C$ establishes
a link to part $P$ for interface $I$, the plugin-generated
\code{\$linkPartToSelf} sets $\$\mathit{selves}_P[I] \leftarrow C$ and
recurses into $P$'s own links, propagating $C$'s identity for each interface
$C$ claims and leaving all other slots unchanged. Each interface of a part resolves
through a single \code{\$selves} slot. Wiring an outer composite that claims
$I$ re-roots this slot to that composite and leaves the slots for interfaces it
does not claim unchanged, so the slot holds exactly one root at any time and
stabilizes once the outermost composite claiming $I$ is wired; $C$ denotes that
final root. Each write occurs during the wiring composite's construction. Wiring happens solely
as link field assignment, and the plugin enforces \code{final} on link
fields, so writes are confined to construction: the \code{\$selves} slots are never
written after the outermost composite claiming $I$ is constructed.

Visibility follows from safe publication: once the composite root is safely published, the
Java Memory Model guarantees that any thread observing it sees all writes that
happened-before publication, the wired slots included. The role of \code{final}
here is write-confinement, not the JMM's final-field freeze: array-element writes
are not covered by freeze semantics, so the guarantee rests on write-confinement
together with safe publication of the root, not on the finality of the slots.

Self-calls in $P$ where $m \in I$ are rewritten by
the plugin to \code{((I)\,\$selves[IDX\_I]).m()}; when $m$ belongs to multiple
interfaces (the diamond case), $I$ is determined at compile time by diamond
resolution (Section~\ref{sec:impl}). For self-calls through the delegation surface, there is no code path in a
part method that dispatches through anything other than the indexed
slot. Because the slot holds $C$ once the composition is constructed and is not
written thereafter, the invariant holds at every point a self-call can execute.
Before wiring, each slot is initialized to \code{this}; an unwired part
dispatches on itself, consistent with standard object-oriented dispatch, and
the invariant is vacuously satisfied (no delegation chain exists).
A full mechanized proof in a Featherweight Java subset is a natural direction
for follow-on work (Section~\ref{sec:futurework}).

%% ─────────────────────────────────────────────────────────────────────────────
\section{Implementation}
\label{sec:impl}
%% ─────────────────────────────────────────────────────────────────────────────

The implementation is a \texttt{javac} plugin operating on the compiler's AST
during two phases. In \textsc{enter}, the plugin generates structural
scaffolding: the \code{\$selves} composite identity array and
\code{\$linkPartToSelf} wiring method for each part, delegation
forwarding method for each composite class, and (for abstract parts) the
\code{\$Impl} subclass and \code{asLink()} factories. In \textsc{analyze},
after type resolution, self-calls in part methods are rewritten,
link field assignments are rewritten to invoke wiring, and
\code{@internal} visibility is enforced. No bytecode manipulation is performed;
all transformation is at the AST level and subject to standard type checking.

\subsection{Part Transformation}

Each part holds a \code{\$selves} array whose compile-time-determined ordinals
cover the part's interfaces.

\begin{lstlisting}[language={},style=java]
private final Object[] $selves = {this, ..., this};
// after wiring: $selves[IDX_I] = composite identity for interface I
\end{lstlisting}

Before wiring, every \code{\$selves} slot holds \code{this}, so the part
dispatches on itself. Wired at link assignment via \code{\$linkPartToSelf}, dispatch routes
through the composite defining the link. \code{\$linkPartToSelf} updates each
relevant slot and recurses through link fields in the part whose delegation
surface includes the interface being propagated, with cycle detection at each
assignment.

Within a part method, every call \code{this.m(args)} where
\code{m} is declared in interface $I$ is rewritten to:

\begin{lstlisting}
((I) $selves[IDX_I]).m(args)
\end{lstlisting}

The cast to $I$ selects the dispatch interface explicitly; when diamond
resolution has assigned $m$ to a specific authoritative link, $I$ is that
link's interface, determined at compile time.

The ordinal \code{IDX\_I} is assigned per part over its own
delegation surface, so it is local to that class: the rewritten self-calls and
the generated \code{\$linkPartToSelf} both reside in the part and share its
indexing, and no index agreement across separately-compiled classes is required.

\subsection{Composite Transformation}

A composite implements every interface claimed by its links. For each method of
a claimed interface that the composite does not implement directly, the plugin
generates a delegating method that forwards through the corresponding
link field:

\begin{lstlisting}
R m(args) { return link.m(args); }
\end{lstlisting}

When two links share an interface, the forwarders for the
shared methods route through the authoritative link
(Definition~\ref{def:diamond}). Wiring corresponding with the authoritative
link is established by sharing the link's part instance between the links
sharing the interface.

In composite classes, each link field assignment \code{this.link = expr}
is rewritten to invoke \code{\$linkPartToSelf}, propagating the composite
identity into the part. An \code{instanceof} check guards the call so
that non-part assignments pass through unchanged.

\subsection{Abstract Part Support}

A part may be abstract, deferring some interface methods to the
composite. Linking an abstract part obligates the composite to implement those
methods or to be declared abstract, enforced at compile time: the plugin
generates a delegating method for every interface method the part implements and
none for a method it leaves abstract, so a concrete composite is missing exactly
those methods and \texttt{javac} rejects it unless they are supplied directly.
This is the compositional analog of the rule that a concrete class must implement
its inherited abstract methods.

To make an abstract part instantiable for linking, the compiler generates a
private \code{\$Impl} subclass whose stubs throw \code{DelegationLinkageError},
and an \code{asLink()} factory per non-private constructor:

\begin{lstlisting}
private static final class $Impl extends Foo {
    $Impl(/* ctor args */) { super(/* ctor args */); }
    @Override public T m(Params p) {
        throw new DelegationLinkageError(
            "abstract method called on an unlinked part");
    }
}
public static Foo asLink(/* ctor args */) { return new $Impl(/* same */); }
\end{lstlisting}

Once wired, the stubs are unreachable: self-calls dispatch through
\code{\$selves} to the composite, whose implementation the compile-time
obligation guarantees. The stubs remain only as a runtime guard for an
\code{asLink()} instance invoked before it is wired.

\subsection{Default Methods}
\label{subsec:default}

A part inherits self-calling behavior from the default methods of the
interfaces it implements. If neither the composite nor the part
overrides a default, it runs with the part as receiver, not the
composite, resulting in incorrect self-call dispatch.

No bytecode instruction directly invokes a default with a receiver other
than the current \code{this}, but a method handle bound to an \code{invokedynamic}
call site can. When a part does not override a default, the plugin
generates an override that invokes the interface's default implementation through
such a site, with the receiver taken from the interface's slot in \code{\$selves},
so the default's self-calls dispatch through that receiver. An explicit
\code{I.super.m()} in part code is rewritten the same way.

\begin{lstlisting}
interface Foo {
    void bar();
    // default self-calls bar()
    default void foo(P p) { bar(); }
}

@part class FooPart implements Foo {
    public void bar() { ... }

    // generated
    // calls Foo.super.foo() with composite in $selves as receiver
    // Foo's bar() self-call dispatches through the composite
    public void foo(P p) {
        invokedynamic Foo::foo((Foo) $selves[IDX_Foo], p);
    }
}
\end{lstlisting}

The site's bootstrap links it once to a fixed target (the default implementation),
with the receiver as its only varying argument. Being
monomorphic, it is inlined by the JIT to the same type of array-indexed
invocation as in Section~\ref{sec:eval}.

\subsection{Safety Guarantees}

\paragraph{Cycle detection.}
If a part is wired to itself directly or transitively,
\code{De\-le\-ga\-tion\-Link\-age\-Er\-ror} is thrown at assignment time during
the recursive \code{\$linkPartToSelf()} call.

\paragraph{Deferred method access.}
The obligation that a wired composite implement an abstract part's deferred methods is enforced at compile time; an \code{asLink()} instance invoked before wiring throws \code{DelegationLinkageError} from the \code{\$Impl} stub, the compositional analog of \code{AbstractMethodError}.

\paragraph{\code{@internal} enforcement.}
\code{@internal} may annotate an interface method or an interface type literal
in a class's \code{implements} clause, with distinct effects. A marked method has its calls confined to the composite, the intent behind \code{protected} in implementation inheritance, but more cleanly scoped for composition. A marked
interface type literal in a class's \code{implements}
clause removes that interface from the class's delegation surface: no composite may claim it via a link field.

\paragraph{Self-preservation.}
Inside a part, \code{this} use is guarded. Field access, private calls,
and callbacks through the part's own link fields (the compositional
analog of \code{super}) are all permitted. Passing or assigning \code{this} as
the concrete part type is not: that is a compile error, since a caller
holding the bare part could invoke methods outside the composite dispatch
path and violate Invariant~\ref{prop:dispatch}. The only \code{this} that can
leave is therefore interface-typed, and the reference that crosses is the
\code{\$selves} slot for that interface: the part's composite identity for
it, which may be the part itself. A client receiving it reaches the
composite's overrides; recovering some other interface by reflection or
downcast can still reach a lower root (Section~\ref{sec:limitations}).

%% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation}
\label{sec:eval}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Dispatch Performance}

The dispatch path for a self-call through a composite is:
\begin{enumerate}
  \item Load \code{\$selves[IDX\_I]}, an array-indexed load with compile-time
        constant index
  \item Cast to $I$
  \item \code{invokeinterface m}, standard JVM virtual dispatch
\end{enumerate}

This is O(1). The parallel to vtable dispatch is direct: a vtable
call is an indexed slot load followed by an indirect call. Call sites in typical
compositions are monomorphic or bimorphic, the JIT's most aggressively optimized
scenarios. Composite dispatch adds at most one additional virtual dispatch over
direct delegation.

\paragraph{Benchmark.}
We verify this empirically using JMH~\cite{jmh} in average-time mode.
Each scenario calls \code{compute()} through a composite root at chain
depth $d$ (1--6). The \emph{baseline} traverses $d$ \code{@link} delegation
hops to a leaf that returns directly, with no self-call. The \emph{concrete}
variant traverses the same hops but the leaf issues a self-call
(\code{\$selves[IDX\_I].scale()}) from a concrete part method,
dispatching through the composite root. The \emph{default} variant issues the
same self-call from an inherited interface default method the leaf does not
override, routed through the generated \code{invokedynamic} override of
Section~\ref{subsec:default}. The delta (concrete minus baseline) isolates the cost of the self-call
dispatch mechanism alone. All runs: 10 JVM forks,
5 warmup iterations, 10 measurement iterations per fork (100 data points per
cell); errors are $\pm$\,half-width of the 99.9\% confidence interval.
A separate single-dispatch calibration run (\code{invokeinterface}
on a monomorphic call site, same harness) yields 0.43\,ns/op; the delta
values in Table~\ref{tab:bench} are therefore roughly 1--2$\times$ a single virtual
dispatch.

\paragraph{Choice of baseline.}
The delta is measured against Parts's own forwarding path, which is the right
baseline: that path is semantically equivalent to Kotlin \code{by} and Lombok
\code{@Delegate} (Section~\ref{sec:related}), so the delta prices exactly the
self-call routing those mechanisms cannot express, against their own dispatch
cost. To our knowledge no other model satisfies all three criteria with a
runnable implementation, so there is no cross-system benchmark to run; that
comparison is analytic (Section~\ref{sec:related}).

\begin{table}[htbp]
\centering
\caption{Self-call dispatch cost by delegation depth (JMH, average-time mode; ns/op, $\pm$\,99.9\% CI half-width). \emph{Baseline} traverses the delegation hops with no self-call; \emph{Concrete} and \emph{Default} add one self-call through the composite, issued from a concrete part method and from an inherited interface default (via \code{invokedynamic}) respectively. The delta (Concrete minus Baseline) isolates the self-call dispatch cost over plain delegation. Values are rounded from the measured means and intervals; the delta combines the Baseline and Concrete cells (difference of means, half-widths in quadrature) and may differ from the rounded column difference by~0.01\,ns.}
\label{tab:bench}
{\small
\begin{tabular}{crrrr}
\toprule
Depth & Baseline (ns) & Concrete (ns) & Default (ns) & Delta (ns) \\
\midrule
1 & $0.62 \pm 0.02$ & $1.16 \pm 0.03$ & $1.18 \pm 0.03$ & $0.55 \pm 0.03$ \\
2 & $0.91 \pm 0.03$ & $1.56 \pm 0.04$ & $1.55 \pm 0.04$ & $0.65 \pm 0.05$ \\
3 & $1.18 \pm 0.03$ & $1.93 \pm 0.05$ & $1.92 \pm 0.05$ & $0.75 \pm 0.06$ \\
4 & $1.64 \pm 0.03$ & $2.30 \pm 0.06$ & $2.23 \pm 0.06$ & $0.66 \pm 0.07$ \\
5 & $1.93 \pm 0.06$ & $2.70 \pm 0.06$ & $2.76 \pm 0.07$ & $0.77 \pm 0.09$ \\
6 & $2.37 \pm 0.06$ & $3.21 \pm 0.09$ & $3.23 \pm 0.08$ & $0.84 \pm 0.11$ \\
\bottomrule
\end{tabular}
}
\end{table}

\noindent
Across depths 1--6 the delta remains bounded, on the order of a single virtual
dispatch (the 0.43\,ns calibration above), consistent with one array-indexed load
(\code{\$selves[IDX\_I]}) followed by \code{invokeinterface}. The self-call routes
through a dynamically wired \code{\$selves} slot, so its receiver is not statically
known and the dispatch does not fold away: the delta is one genuine virtual
dispatch, with the array indexing on top (confirmed free below). The O(1) claim is
structural: one dispatch per self-call, independent of depth, with the per-call
cost pinned by the isolation measurement below. The small upward drift in the
delta across depths (0.55 to 0.84\,ns) stays well under a second dispatch and
plausibly reflects less complete inlining of the dispatch site within the larger
compiled method at greater depth. The \emph{default} column tracks the \emph{concrete}
column within $\pm$3\% at every depth, with the sign of the difference varying;
the two are indistinguishable at this resolution. The interface-default template,
dispatched through \code{invokedynamic}, thus carries no measurable cost over the
concrete part template, confirming the equivalence claimed in
Section~\ref{subsec:default}. The comparison is steady-state: the one-time
bootstrap that links each \code{invokedynamic} site is absorbed during warmup and
is not measured here.

\paragraph{Isolating the array index.}
The depth-series delta is essentially one virtual dispatch; to confirm the
\code{\$selves} array indexing adds nothing on top, we isolate the dispatch.
Holding a target method non-inlinable so a real call occurs, we compare three
depth-0 forms over the same receiver: a vtable call (\code{invokevirtual}), the
\code{invokeinterface} it compiles to, and the plugin's emitted form
\code{((I)\,\$selves[IDX\_I]).m()} (an \code{Object[]} element load, a cast, and
\code{invokeinterface}). Over 10 JVM forks (JDK~21) the three are statistically
indistinguishable: the ISD form differs from the bare \code{invokeinterface} it
emits by $0.03\pm0.09$\,ns, and from \code{invokevirtual} by $0.10\pm0.05$\,ns,
both within the fork-to-fork variance of the measurement. The array-element load
is thus not separately observable; it overlaps with the dispatch. The self-call
costs one virtual dispatch and no more, the empirical content of the
vtable-equivalence claim.

\paragraph{Construction cost.}
The result above is the steady-state call cost; wiring is paid once, at
construction. Building a composite invokes \code{\$linkPartToSelf} at each link
assignment, which sets the relevant \code{\$selves} slots and recurses through the
part's own link fields, propagating composite identity for each interface the
composite claims. Cycle detection adds only a constant-time check per claimed
interface, an \code{aaload} and a compare, so each part node costs O(1) per
propagated interface and the whole traversal is linear in the size of the
delegation graph reachable through link fields. None of it lies on the dispatch
path, so the per-call cost is unchanged, and the traversal is proportional to the
object graph the program constructs in any case. The cost is therefore a bounded,
one-time increment to composite construction, amortized over every subsequent
self-call.

\subsection{Example: Teaching Assistant}

The following example exercises DAG composition, diamond dispatch resolution,
and self-call routing through the composite. A teaching assistant is both
a student and a teacher; both roles share a common \code{Person} base.

\begin{lstlisting}
interface Person {
    String getName(); String getTitle(); String getTitledName();
}
interface Student extends Person { String getMajor(); }
interface Teacher extends Person { String getDept();   }
interface TA extends Student, Teacher {}
\end{lstlisting}

\paragraph{Forwarding (Kotlin)~\cite{kotlin,lombok}.}
Forwarding-based mechanisms inject the implementation as a real object at
construction time; components are independent. But self-calls remain within
the component: the composite is invisible. Delegating
\code{Student by s} and \code{Teacher by t} creates a diamond: the compiler
forces a manual override of every shared \code{Person} method. The deeper problem is that
\code{getTitledName()}, the template method worth sharing, cannot be safely
delegated: whichever branch receives it calls \code{getTitle()} on its own
\code{this}, bypassing \code{TaImpl.getTitle()}.

\begin{lstlisting}[style=kotlin]
class TaImpl(val s: Student, val t: Teacher) : TA,
    Student by s, Teacher by t {
  // Diamond: every shared Person method must be manually resolved
  override fun getName()  = s.getName()
  override fun getTitle() = "TA"
  // Without this override, getTitledName() forwards to s.getTitledName(),
  // which calls s.getTitle() = "Student" -- self stays in the delegate:
  //   TaImpl(...).getTitledName()  ->  "Student Milton"  (wrong)
  override fun getTitledName() = "${getTitle()} ${getName()}"
  //   TaImpl(...).getTitledName()  ->  "TA Milton"  (correct, but forced)
}
\end{lstlisting}

Every composite must re-implement \code{getTitledName()}.
Adding a new template method to \code{Person} requires a matching override in
every composite class; the maintenance cost grows with the number of composites. Lombok's
\code{@Delegate} generates equivalent forwarders in Java; the self-call
semantics are identical.

\paragraph{Parts.}
\code{PersonPart} holds the template once. Its self-call to \code{getTitle()}
dispatches through the composite at each level of the delegation chain.

\begin{lstlisting}
@part class PersonPart implements Person {
  private final String name;
  public PersonPart(String name) { this.name = name; }
  public String getName()       { return name; }
  public String getTitle()      { return "Person"; }
  public String getTitledName() { return getTitle() + " " + getName(); }
}
@part class StudentPart implements Student {
  @link Person person;
  private final String major;
  public StudentPart(Person p, String major) {
    this.person = p; this.major = major;
  }
  public String getTitle()  { return "Student"; }
  public String getMajor()  { return major; }
}
@part class TeacherPart implements Teacher {
  @link Person person;
  private final String dept;
  public TeacherPart(Person p, String dept) {
    this.person = p; this.dept = dept;
  }
  public String getTitle() { return "Teacher"; }
  public String getDept()  { return dept; }
}
@part class TaPart implements TA {
  @link(share=Person.class) Student student;
  @link Teacher teacher;
  public TaPart(Student s) {
    this.student = s;
    this.teacher = new TeacherPart(s, "Math");
  }
  public String getTitle() { return "TA"; }
}

TA ta = new TaPart(new StudentPart(new PersonPart("Milton"), "CS"));
ta.getTitledName(); // "TA Milton"  -- getTitle() dispatches through TaPart
\end{lstlisting}

\code{getTitledName()} is defined once in \code{PersonPart} and overridden
nowhere. Diamond resolution is handled statically: \code{@link(share=Person.class)}
designates \code{student} as the authoritative delegate for \code{Person},
so \code{Person} is excluded from propagation when \code{teacher} is wired,
leaving the identity established through \code{student} intact.

\subsection{Case Study: Stream Decoration}

A pattern common in the JDK, exemplified by \code{java.io.InputStream}~\cite{gamma1994}, is
a bulk operation implemented as a loop over a finer one: \code{read(byte[],
int, int)} calls the single-byte \code{read()} repeatedly. Decorators override
\code{read()} to add behavior; the template is supposed to propagate it to bulk
calls automatically. The \code{Readable} interface below captures this contract
directly. Under forwarding, the propagation breaks.

A \code{CountingStream} that tracks bytes consumed overrides \code{read()} to
increment a counter. With Lombok's \code{@Delegate}, \code{read(byte[], int, int)} is
forwarded to the delegate, which calls \code{read()} on itself, not on
\code{CountingStream}. The override is never reached for bulk calls.

\paragraph{Forwarding~\cite{lombok}.}\leavevmode

\begin{lstlisting}
interface Readable {
    int read();
    int read(byte[] buf, int off, int len);
}

class BaseStream implements Readable {
    private final byte[] data;
    private int pos;
    BaseStream(byte[] data) { this.data = data; }
    public int read() {
        return pos < data.length ? (data[pos++] & 0xff) : -1;
    }
    public int read(byte[] buf, int off, int len) {
        for (int i = 0; i < len; i++) {
            int b = read();                      // self-call
            if (b == -1) return i == 0 ? -1 : i;
            buf[off + i] = (byte) b;
        }
        return len;
    }
}

class CountingStream implements Readable {
    @Delegate private final Readable delegate;
    private int count;
    CountingStream(Readable r) { this.delegate = r; }
    @Override public int read() {
        int b = delegate.read();
        if (b != -1) count++;
        return b;
    }
    public int count() { return count; }
}

CountingStream cs = new CountingStream(new BaseStream(data));
cs.read(buf, 0, 16);  // -> delegate.read(byte[],int,int)
                       //    -> delegate.read()  -- not cs.read()
cs.count();            // 0  -- bulk reads not counted
\end{lstlisting}

\paragraph{Parts.}
\code{BaseStream} is unchanged except for the \code{@part} annotation. The
self-call in \code{read(byte[], int, int)} routes through \code{CountingStream}
on every call path.

\begin{lstlisting}
@part class BaseStream implements Readable { /* same as above */ }

class CountingStream implements Readable {
    @link Readable stream;
    private int count;
    CountingStream(Readable r) { this.stream = r; }
    @Override public int read() {
        int b = stream.read();
        if (b != -1) count++;
        return b;
    }
    public int count() { return count; }
}

CountingStream cs = new CountingStream(new BaseStream(data));
cs.read(buf, 0, 16);  // -> BaseStream.read(byte[],int,int)
                       //    -> self-call read() -> CountingStream.read()
cs.count();            // 16  -- all bytes counted
\end{lstlisting}

A new decorator (buffering, decryption, rate limiting) requires only
overriding \code{read()}; \code{BaseStream.read(byte[], int, int)} routes its
self-call through whichever composite is the dispatch root, without
modification.

\subsection{Limitations}\label{sec:limitations}

\begin{itemize}
  \item Using \code{@link}, \code{@part}, and \code{@internal} requires
        configuring the \texttt{javac} plugin in the build system (Maven,
        Gradle, or similar).
  \item Composite dispatch semantics are not enforced against runtime
        introspection that reaches past a reference's supplied static type. A
        client that reflects on an escaped reference, or downcasts it to an
        interface it was not handed, can invoke that interface below its
        composite root, bypassing the overrides above it; type-directed
        clients, which use the interface they were handed, are unaffected.
  \item A part instance carries one composite root per interface
        (Section~\ref{sec:discussion}). Linking the same instance into two
        composites that each claim the \emph{same} interface, with neither
        nesting the other, re-roots that interface to whichever is wired last,
        redirecting the first composite's self-calls. The contention
        is undetectable at the wiring step: re-rooting an already-wired slot
        is the model's normal operation during transitive propagation, so the
        two cases are observationally identical. One instance cannot be the
        composite root for a single interface in two places at once; separate
        part instances avoid it entirely.
  \item The plugin relies on internal \texttt{javac} APIs for AST
        rewriting, which may require adaptation across JDK releases.
        In practice the approach has proven stable: the Manifold project,
        on which this plugin is built, has tracked 18 JDK releases
        without API-driven breakage.
\end{itemize}

%% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}
%% ─────────────────────────────────────────────────────────────────────────────

\subsection{Implementation Identity and Dispatch Identity}

A composition model that supports independent components must handle two
structural scenarios:

\begin{itemize}
  \item \textbf{Partial delegation}: a composite delegates a subset of a
        part's interfaces.
  \item \textbf{Multi-root}: a part's interfaces are claimed by separate
        composites, each requiring a distinct dispatch root from the same
        instance.
\end{itemize}

Multi-root requires a part to carry multiple simultaneous
composite identities, one per interface. ISD resolves this by separating two
notions of identity: \emph{implementation identity}, the part object, from
\emph{dispatch identity}, the composite per interface. Within ISD, partial delegation and multi-root are modeled the same way. Every
interface has a single composite root: the composite that claims it, or the
part itself when none does.

Multi-root arises naturally from partial delegation: when an outer composite claims a subset of a
part's interfaces, transitive propagation assigns different roots to the
remaining interfaces, producing multi-root at the part level. The \code{Track} example in
Section~\ref{sec:model} demonstrates this directly: \code{Library} claims
only \code{Classifiable} from \code{Playlist} (partial delegation), which by propagation
gives \code{Track} separate roots for \code{Streamable} and
\code{Classifiable} (multi-root).

\subsection{Language Design Implications}

Parts resolves each known deficiency of implementation inheritance:

\begin{itemize}
  \item \textbf{Fragile base class}~\cite{mikhajlov1998}: a composite's dependency surface is
        the interface contracts it explicitly claims, not a part's internal
        implementation; changes to its encapsulated internals do not propagate. Partial
        delegation completes the isolation: a composite is unaffected by changes
        to interfaces it did not claim.
  \item \textbf{Forced IS-A semantics}: reuse does not require a type
        relationship; the composite is a subtype of the interface, not of the part.
  \item \textbf{The \emph{self} problem}: a hazard in forwarding-based composition
        where self-calls remain within the component rather than routing
        through the composite; resolved by ISD.
  \item \textbf{Diamond ambiguity}: explicit \code{@link(share=T.class)} resolution;
        unresolved diamonds are compile errors, not silent merges.
  \item \textbf{Encapsulation breach}: a composite may implement a subset
        of its parts' interfaces; those it does not implement are absent from
        its API by construction. \code{@internal} strengthens this further:
        it provides composite-scoped visibility, preventing designated
        interfaces from being claimed by any composite.
  \item \textbf{Abstract behavior}: abstract parts with deferred
        linkage (\code{asLink()}) preserve the template-method capability
        of abstract classes without coupling the hierarchy.
\end{itemize}

Implementation inheritance's power is open recursion. Its fragility is the surface that power ranges over: every overridable member. Parts keeps the open recursion but bounds the surface to the declared interface contract; self-calls still reach the composite's overrides, while the internals that make base classes fragile stay off it entirely.

The interface contract---not implementation types---is what makes dispatch constant-time: a fixed, statically-known interface set lets each interface's composite root be an array-indexed slot. Restricting a part from exposing \code{this} as its concrete type is the static guarantee that makes the dispatch invariant hold.

Implementation inheritance and Parts are two bindings of the same open
recursion: implementation inheritance binds it to the class hierarchy, where a
subclass IS-A its superclass and inherits its entire implementation surface,
while Parts binds it to a held object and a declared interface, where the
composite HAS-A its parts and IS-A only their interfaces. That inheritance
is open recursion, separable from subtyping, is
Cook's~\cite{cookpalsberg1989,cook1990}, and the reading of inheritance as
composition is Bracha and Cook's~\cite{bracha1990}. Cook's model threads self as
an explicit parameter; ISD keeps it implicit and per-interface. The missing piece
was not theoretical but practical: independent runtime components, static interface
contracts, and dispatch at the cost of a virtual call.

\subsection{Concurrency}

A constructed composition is safe to share across threads. Each interface's
\code{\$selves} slot is written only while its final composite root
(Invariant~\ref{prop:dispatch}) is under construction; once that root is safely
published the slot is never written again, so the JMM makes all wired state
visible to any thread observing the root, with no synchronization on the dispatch
path. The guarantee is thus scoped to a single composition, resting on the
one-root invariant of Section~\ref{sec:discussion}. Parts may hold mutable
state under the usual Java rules; dispatch adds no synchronization points.

\subsection{Future Work}
\label{sec:futurework}

\paragraph{Per-interface composite roots as a general primitive.}
Assigning dispatch identity per interface rather than per object is a primitive
that prior composition models do not expose. Problems that current statically-typed designs address through dynamic dispatch, AOP weaving, or collapsed component hierarchies may have cleaner solutions within this primitive. The extent of that solution space is an open question.

\paragraph{Formal verification.}
A mechanized proof of Invariant~\ref{prop:dispatch} in a small-step operational
semantics would provide assurance beyond the construction argument in Section~\ref{sec:model}. The
model maps cleanly to a subset of Featherweight Java, a natural starting point.

\paragraph{Domain model and persistence integration.}
IS-A relationships in normalized relational schemas map naturally to the
Parts model: each type owns its interface and its data independently,
composed at construction with open recursion available across the composed
object. Transactional systems are designed for this normalized, per-type
form. Much of the pressure to denormalize comes from the object model: the
table-per-hierarchy strategy---a common ORM default for IS-A---collapses the
hierarchy into one table, the storage shadow of the single object
implementation inheritance produces. Mapping such a schema to objects has
meant a choice between an inheritance hierarchy that couples the types and
hand-wired composition that gives up open recursion. Parts needs neither,
keeping the types independent with open recursion intact, aligning the domain
model with the normalized schema the relational model already favors. How that
mapping is realized in a persistence framework is an open question.

\paragraph{Parts as a complete compositional alternative.}
Parts does not argue that implementation inheritance is wrong, but that
it is no longer the only practical option. Any codebase written with
implementation inheritance can in principle be expressed purely in Parts
compositions; whether the result is better depends on the codebase. What
the model offers is a compositional foundation that delivers open recursion
without hierarchy and fosters low coupling and high cohesion through its
interface contracts and independent parts. A systematic study of implementation
inheritance patterns in real codebases would identify any remaining gaps
and clarify where the compositional model offers the most advantage.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}
%% ─────────────────────────────────────────────────────────────────────────────

Static typing is the baseline requirement; models without it are included where relevant.

\paragraph{Forwarding.}
Kotlin's \code{by} keyword~\cite{kotlin} generates forwarding methods at compile
time; Lombok's \code{@Delegate}~\cite{lombok} does the same in Java; Scala~3's
\code{export} keyword~\cite{scala3export} generates forwarding methods for
delegated members.
In all cases, a component's self-calls resolve within the component; the composite
host is invisible. The mechanism is forwarding, not delegation. Scala
provides both mechanisms in the same language: \code{export}
for forwarding with independent components, and traits for correct self-call
dispatch. Yet no combination of the two satisfies both requirements.
B\"uchi and Weck's generic wrappers~\cite{buchi2000} give the same forwarding
choice a formal, soundness-proven treatment, and are the closest prior analysis of
the identity-escape hazard our self-preservation constraint prevents
(Section~\ref{sec:impl}), though they decline the disciplines we adopt as too
restrictive for arbitrary legacy objects.

\paragraph{Scala traits / abstract type members.}
Odersky and Zenger~\cite{odersky2005} introduced statically-typed mixin
composition with abstract type members. Scala's \code{with} clause linearizes
traits into the target class at compile time, so self-calls dispatch through the
outermost class in the linearized hierarchy. This gives open recursion, but it
also fuses the component into the class: the trait's implementation is resolved
at compile time, not supplied as an independent component at construction. A component implementation therefore cannot be chosen or replaced, and
a single instance cannot be shared across composites. The fusion and the open recursion are inseparable: linearization is the single
mechanism behind both, so in statically-typed languages fusing the component
into the class became standard. Interface-scoped dispatch shows the fusion
was never necessary: it obtains the same open recursion without abandoning independent components.

\paragraph{Prototype-based delegation.}
Lieberman~\cite{lieberman1986} introduced \emph{open delegation}: an object may
delegate unanswered messages to a prototype, with \emph{self} remaining the
original receiver throughout the chain. Self~\cite{self1987} realized this in a
working language. These systems support open recursion with late-bound
components in arbitrary delegation graphs. Static type guarantees are absent by
design.

\paragraph{Mixins.}
Bracha and Cook~\cite{bracha1990} formalized mixins as parameterized abstract
subclasses. Mixins extend through linearization: the mixin is applied to a chain at class
definition time, not supplied as an independent component at construction.
Self-calls dispatch through \code{this} in the linearized result.

\paragraph{Traits.}
Sch\"arli et al.~\cite{scharli2003} proposed traits as stateless, composable
units of behavior. Trait composition is a compile-time operation: methods are
flattened into the target class, not supplied as independent components at
construction. Self-calls dispatch through \code{this} in the composed class.

\paragraph{Multiple inheritance.}
Stroustrup~\cite{stroustrup1987} introduced virtual base classes and virtual
inheritance in C++ to handle diamonds. The result is a single linearized object
with a merged vtable; components are not preserved as independent objects.

\paragraph{Kniesel / Lava.}
Kniesel's ``type-safe delegation''~\cite{kniesel1999,kniesel2000} is the closest
predecessor and frames the problem this paper inherits: components are real
runtime objects supplied at construction, types are statically enforced, and
DARWIN binds \texttt{self} to the original receiver so self-calls route through
the composite rather than staying in the component. The difference from Parts is
not in capability but in practice. DARWIN routes each self-call through forwarding adapters and a reflective walk
up the parent-pointer chain (\code{Lava.parentCast}); that walk is a search, not
a constant-time step. The typical configuration \code{@link} targets is an abstract or interface
parent with a concrete-subtype instance. For exactly that case, Kniesel's own
analysis is candid that dispatch ``can be slowed down
significantly''~\cite[\S7.4.2]{kniesel2000}. The component is never
present in the composition as itself---widened to \code{Object} or
adapter-wrapped---where Parts holds it directly and dispatches through an array
index. He explored mitigations, leaving their evaluation
open~\cite[\S7.5.5]{kniesel2000}; the system was realized in Lava, a research
language. Kniesel also weighed the static
alternative of storing the composite identity in the component, and his objections
to it are sound: ``sharing of one parent by multiple delegating children
cannot be expressed at all,'' and recursive delegation could be simulated only
``with significant run-time and software maintenance costs''~\cite{kniesel1999}.
Each objection holds for a single per-object identity wired at run time.
Interface-scoped dispatch answers them with two mechanisms:
indexing the stored identity by interface lets multiple composites claiming disjoint
interfaces of a shared part each hold their own root, and transitive propagation
wires recursive delegation chains at construction, so dispatch stays constant-time
and the chain adds no per-call cost or maintenance burden.

\paragraph{Delegation Layers.}
Ostermann's delegation layers~\cite{ostermann2002} (ECOOP 2002) combine
delegation with virtual classes to compose collaborations dynamically at
runtime; self-calls route correctly within a composed collaboration. The unit
composed is the collaboration (a layer of virtual nested classes bound to a
common enclosing type), not the individual object; participants are members of
that collaboration.

\paragraph{Virtual classes and family polymorphism.}
The virtual-class lineage (gbeta~\cite{ernst2001}, CaesarJ~\cite{aracic2006},
Object Teams/Java~\cite{herrmann2002}, and J\&~\cite{nystrom2006}) achieves open class composition through virtual nested
classes within a common enclosing family type; Ernst, Ostermann, and
Cook~\cite{ernst2006} provide the definitive formal treatment.
Components are runtime instances of virtual classes, but they are bound to an
enclosing family type and cannot be independently supplied from outside.
Self-calls dispatch correctly within the family. The dispatch scope is the
enclosing family.

\paragraph{Dynamic composition.}
Newspeak~\cite{bracha2010} treats classes as first-class late-bound arguments,
enabling runtime composition of first-class module objects with open
recursion. Static type checking is absent by design.

%% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
%% ─────────────────────────────────────────────────────────────────────────────

In static typing, the \emph{self} problem with independent components is a
large part of why object composition never displaced implementation inheritance:
a self-call that reaches the composite's overrides is the one capability that
made inheritance necessary rather than merely convenient. Parts resolves this by grounding identity in the interface rather than the
object, recovering open recursion across independent components at ordinary
dispatch cost---the two properties are not inherently in conflict, only
unreconciled by prior models.


%% ─────────────────────────────────────────────────────────────────────────────
%% References
%% ─────────────────────────────────────────────────────────────────────────────

\begin{thebibliography}{10}

\bibitem{gamma1994}
E.~Gamma, R.~Helm, R.~Johnson, and J.~Vlissides.
\newblock {\em Design Patterns: Elements of Reusable Object-Oriented Software}.
\newblock Addison-Wesley, 1994.

\bibitem{mikhajlov1998}
L.~Mikhajlov and E.~Sekerinski.
\newblock A study of the fragile base class problem.
\newblock In {\em Proceedings of ECOOP}, LNCS 1445, pages 355--382, 1998.

\bibitem{lieberman1986}
H.~Lieberman.
\newblock Using prototypical objects to implement shared behavior in
  object-oriented systems.
\newblock In {\em Proceedings of OOPSLA}, pages 214--223, 1986.

\bibitem{bracha1990}
G.~Bracha and W.~Cook.
\newblock Mixin-based inheritance.
\newblock In {\em Proceedings of OOPSLA/ECOOP}, pages 303--311, 1990.

\bibitem{cookpalsberg1989}
W.~Cook and J.~Palsberg.
\newblock A denotational semantics of inheritance and its correctness.
\newblock In {\em Proceedings of OOPSLA}, pages 433--443, 1989.

\bibitem{cook1990}
W.~R.~Cook, W.~Hill, and P.~S.~Canning.
\newblock Inheritance is not subtyping.
\newblock In {\em Conference Record of POPL}, pages 125--135, 1990.

\bibitem{scharli2003}
N.~Sch{\"a}rli, S.~Ducasse, O.~Nierstrasz, and A.~Black.
\newblock Traits: Composable units of behaviour.
\newblock In {\em Proceedings of ECOOP}, LNCS 2743, pages 248--274, 2003.

\bibitem{self1987}
D.~Ungar and R.~B.~Smith.
\newblock Self: The power of simplicity.
\newblock In {\em Proceedings of OOPSLA}, pages 227--242, 1987.

\bibitem{odersky2005}
M.~Odersky and M.~Zenger.
\newblock Scalable component abstractions.
\newblock In {\em Proceedings of OOPSLA}, pages 41--57, 2005.

\bibitem{stroustrup1987}
B.~Stroustrup.
\newblock Multiple inheritance for {C++}.
\newblock In {\em Proceedings of the EUUG Spring Conference}, 1987.

\bibitem{kniesel1999}
G.~Kniesel.
\newblock Type-safe delegation for run-time component adaptation.
\newblock In {\em Proceedings of ECOOP}, LNCS 1628, pages 351--366, 1999.
\newblock DOI: \href{https://doi.org/10.1007/3-540-48743-3_16}{10.1007/3-540-48743-3\_16}.

\bibitem{kniesel2000}
G.~Kniesel.
\newblock {\em Dynamic Object-Based Inheritance with Subtyping}.
\newblock PhD thesis, University of Bonn, 2000.

\bibitem{buchi2000}
M.~B\"uchi and W.~Weck.
\newblock Generic wrappers.
\newblock In {\em Proceedings of ECOOP}, LNCS 1850, pages 201--225, 2000.

\bibitem{ostermann2002}
K.~Ostermann.
\newblock Dynamically composable collaborations with delegation layers.
\newblock In {\em Proceedings of ECOOP}, LNCS 2374, pages 89--110, 2002.

\bibitem{ernst2006}
E.~Ernst, K.~Ostermann, and W.~R.~Cook.
\newblock A virtual class calculus.
\newblock In {\em Conference Record of POPL}, pages 270--282, 2006.

\bibitem{nystrom2006}
N.~Nystrom, X.~Qi, and A.~C.~Myers.
\newblock {J\&}: Nested intersection for scalable software composition.
\newblock In {\em Proceedings of OOPSLA}, pages 347--366, 2006.

\bibitem{ernst2001}
E.~Ernst.
\newblock Family polymorphism.
\newblock In {\em Proceedings of ECOOP}, LNCS 2072, pages 303--326, 2001.

\bibitem{aracic2006}
I.~Aracic, V.~Gasiunas, M.~Mezini, and K.~Ostermann.
\newblock An overview of {CaesarJ}.
\newblock In {\em Transactions on Aspect-Oriented Software Development I}, LNCS 3880, pages 135--173, 2006.

\bibitem{herrmann2002}
S.~Herrmann.
\newblock Object {Teams}: Improving modularity for crosscutting collaborations.
\newblock In {\em Proceedings of NetObjectDays}, LNCS 2591, pages 248--264, 2002.

\bibitem{bracha2010}
G.~Bracha, P.~von~der~Ah{\'e}, V.~Bykov, Y.~Kashai, W.~Maddox, and E.~Miranda.
\newblock Modules as objects in {Newspeak}.
\newblock In {\em Proceedings of ECOOP}, LNCS 6183, pages 405--428, 2010.

\bibitem{scala3export}
EPFL.
\newblock Exports --- {Scala~3} reference.
\newblock \url{https://docs.scala-lang.org/scala3/reference/other-new-features/export.html}, 2024.

\bibitem{kotlin}
JetBrains.
\newblock Delegation --- {Kotlin} programming language.
\newblock \url{https://kotlinlang.org/docs/delegation.html}, 2024.

\bibitem{lombok}
Project Lombok.
\newblock \texttt{@Delegate} --- {Lombok} features.
\newblock \url{https://projectlombok.org/features/Delegate}, 2024.

\bibitem{jmh}
A.~Shipilev and contributors.
\newblock {JMH}: Java Microbenchmark Harness.
\newblock \url{https://github.com/openjdk/jmh}, 2013--2024.

\bibitem{mckinney2026}
R.~S.~McKinney.
\newblock manifold-parts.
\newblock \url{https://github.com/manifold-systems/manifold/tree/master/manifold-deps-parent/manifold-parts}, 2026.

\end{thebibliography}





\end{document}
